Faithful Model Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-12-14 paper/camera-ready/altmeyer.tex
resolve overflow issue
text updated

\begin{equation} \label{eq:general}
\begin{aligned}
\mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
\end{aligned} 
\end{equation}

@@

To relax this assumption, we use conformal prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}. Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training. It works under the premise of turning heuristic notions of uncertainty into rigorous estimates by repeatedly sifting through the training data or a dedicated calibration dataset.

@@

The key observation in our context is that we can rely solely on decreasing the energy of the counterfactual itself. This is sufficient to capture the generative property of the underlying model since it is implicitly captured by its parameters $\theta$. Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all as we explain in the appendix.

@@

For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}~\citep{lecun1998gradient}). Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}. To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset. Full details concerning our parameter choices, training procedures and model performance can be found in the appendix.

We perform benchmarks on eight datasets from different domains. From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit}~\citet{hoffman1994german} and \textit{California Housing}~\citet{pace1997sparse}. All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}. Following related literature~\citep{schut2021generating,dhurandhar2018explanations} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}. Detailed descriptions and results for all datasets can be found in the appendix.

In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}. The tables show sample averages along with standard deviations for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}). For each metric, the best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively. The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}. For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.

@@

Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness. Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform all other generators with respect to faithfulness, in many cases substantially. This pattern is mostly robust across all other benchmark datasets (additional results can be found in the appendix).

In particular, we note that the best results are generally obtained when using the full \textit{ECCCo} objective (Equation~\ref{eq:eccco}). In other words, constraining both energy and predictive uncertainty typically yields the most faithful counterfactuals. We expected the former to play a more significant role in this context and that is typically what we find across all datasets. For example, the results for \textit{GMSC} in Table~\ref{tab:results-tabular} indicate that faithfulness can be improved substantially by relying solely on the energy constraint (\textit{ECCCo (no CP)}). In some cases though, as for the \textit{California Housing} dataset, \textit{ECCCo (no EBM)} actually outperforms \textit{ECCCo (no CP)}. This indicates that predictive uncertainty minimization plays an important role in achieving faithfulness.

We also generally find that the highest degree of faithfulness is obtained when the counterfactual search is performed directly in the feature space $\mathcal{X}$. While \textit{ECCCo+} typically attains high levels of faithfulness compared to most other generators, it is consistently outperformed by \textit{ECCCo}. The case is even stronger for \textit{REVISE}, which performs worst out of all generators for faithfulness on the \textit{GMSC} dataset and better only than \textit{Wachter} on \textit{California Housing}.

These findings are consistent with the notion that surrogate models may inhibit faithfulness. Even though dimensionality reduction through PCA in the case of \textit{ECCCo+} can be considered a relatively mild form of intervention, the first $n_z$ principal components fail to capture some of the variation in the data, that the underlying model itself may be sensitive to. This notion is illustrated nicely in Figure~\ref{fig:mnist-bmk}, where the counterfactual produced by \textit{ECCCo} is somewhat noisier and grainier than the one produced by \textit{ECCCo+}.

In conclusion, we recommend in light of the findings here to use the full \textit{ECCCo} search objective whenever model faithfulness is a key priority.

@@

Overall, we find strong empirical evidence suggesting that \textit{ECCCo} can achieve near state-of-the-art plausibility without sacrificing faithfulness. Figure~\ref{fig:mnist-bmk} shows one such example taken from the \textit{MNIST} benchmark where the objective is to turn the factual three (far left) into a five. The underlying model is a LeNet-5 \textit{CNN}. The different images show the counterfactuals produced by the generators, of which all but the one produced by \textit{Schut} are valid. Both variations of \textit{ECCCo} produce plausible counterfactuals.

@@

While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to trade off some degree of faithfulness for plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context at the small cost of slightly reduced faithfulness. For the vision datasets especially, we find that  \textit{ECCCo+} is consistently second only to \textit{REVISE} for all models and regularly substantially better than the baseline. Looking at the \textit{California Housing} data, latent space search markedly improves plausibility without sacrificing faithfulness: for the \textit{JEM} Ensemble, \textit{ECCCo+} performs substantially better than the baseline and only marginally worse than \textit{REVISE}. Importantly, \textit{ECCCo+} does not attain plausibility at all costs: for the MLP, plausibility is still very low but this seems to faithfully represent what the model has learned.

We conclude that \textit{ECCCo} offers us a way to balance the objectives of faithfulness and plausibility. \textit{ECCCo+} can be used to tilt the scale in favour of plausibility if needed.

@@

Another commonly discussed desideratum is closeness~\citep{wachter2017counterfactual}: counterfactuals that are closer to their factuals are associated with smaller costs to individuals in the context of algorithmic recourse. As evident from the additional tables in the appendix, the closeness desideratum tends to be negatively correlated with plausibility and faithfulness. Consequently, both \textit{REVISE} and \textit{ECCCo} generally yield more costly counterfactuals than the baseline. Nonetheless, \textit{ECCCo} does not seem to stretch costs unnecessarily: in Figure~\ref{fig:mnist-bmk} useful parts of the factual three are clearly retained.

>>>>

\begin{equation} \label{eq:general}
\begin{aligned}
\min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) }  \right\} 
\end{aligned} 
\end{equation}

@@

To relax this assumption, we use conformal prediction (CP), an approach to predictive uncertainty quantification that has recently gained popularity~\citep{angelopoulos2021gentle,manokhin2022awesome}. Crucially for our intended application, CP is model-agnostic and can be applied during inference without placing any restrictions on model training. It works under the premise of turning heuristic notions of uncertainty into rigorous estimates by repeatedly sifting through the training data or a dedicated calibration dataset. Calibration data is used to compute so-called nonconformity scores: $\mathcal{S}=\{s(\mathbf{x}_i,\mathbf{y}_i)\}_{i \in \mathcal{D}_{\text{cal}}}$ where $s: (\mathcal{X},\mathcal{Y}) \mapsto \mathbb{R}$ is referred to as \textit{score function} (see appendix for details).

@@

The key observation in our context is that we can rely solely on decreasing the energy of the counterfactual itself. This is sufficient to capture the generative property of the underlying model since it is implicitly captured by its parameters $\theta$. Importantly, this means that we do not need to generate conditional samples through SGLD during our counterfactual search at all (see appendix for details).

@@

For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}~\citep{lecun1998gradient}). Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}. To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.

We perform benchmarks on eight datasets from different domains. From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit}~\citet{hoffman1994german} and \textit{California Housing}~\citet{pace1997sparse}. All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}. Following related literature~\citep{schut2021generating,dhurandhar2018explanations} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.

Full details concerning model training as well as detailed descriptions and results for all datasets can be found in the appendix. In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}. The tables show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}). For each metric, the best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively. The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}. For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.

@@

Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness. Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform other generators with respect to faithfulness, in many cases substantially. This pattern is mostly robust across all other datasets.

In particular, we note that the best results are generally obtained when using the full \textit{ECCCo} objective (Equation~\ref{eq:eccco}). In other words, constraining both energy and predictive uncertainty typically yields the most faithful counterfactuals. We expected the former to play a more significant role in this context and that is typically what we find across all datasets. The results in Table~\ref{tab:results-tabular} indicate that faithfulness can be improved substantially by relying solely on the energy constraint (\textit{ECCCo (no CP)}). In most cases, however, the full objective yields the most faithful counterfactuals. This indicates that predictive uncertainty minimization plays an important role in achieving faithfulness.

We also generally find that latent space search does not impede faithfulness for \textit{ECCCo}. In most cases \textit{ECCCo+} is either on par with \textit{ECCCo} or even outperforms it. There are some notable exceptions though. Cases in which \textit{ECCCo} achieves substantially better faithfulness without latent space search tend to involve more vulnerable models like the simple MLP for MNIST in Table~\ref{tab:results-vision}. We explain this finding as follows: even though dimensionality reduction through PCA in the case of \textit{ECCCo+} can be considered a relatively mild form of intervention, the first $n_z$ principal components fail to capture some of the variation in the data. More vulnerable models may be particularly sensitive to this residual variation in the data.

Consistent with this finding, we also observe that \textit{REVISE} ranks higher for faithfulness, if the model itself has learned more plausible representations of the underlying data: \textit{REVISE} generates more faithful counterfactuals than the baseline for the \textit{JEM} Ensemble in Table~\ref{tab:results-tabular} and the LeNet-5 \textit{CNN} in Table~\ref{tab:results-vision}. This demonstrates that the two desiderata---faithfulness and plausibility---are not mutually exclusive.

@@

Overall, we find strong empirical evidence suggesting that \textit{ECCCo} can achieve near state-of-the-art plausibility without sacrificing faithfulness. Figure~\ref{fig:mnist-bmk} shows one such example taken from the \textit{MNIST} benchmark where the objective is to turn the factual `three' (far left) into a `five'. The underlying model is a LeNet-5 \textit{CNN}. The different images show the counterfactuals produced by the generators, of which all but the one produced by \textit{Schut} are valid. Both variations of \textit{ECCCo} produce plausible counterfactuals.

@@

While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to balance faithfulness and plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context at the small cost of slightly reduced faithfulness. For the vision datasets especially, we find that  \textit{ECCCo+} is consistently second only to \textit{REVISE} for all models and regularly substantially better than the baseline. Looking at the \textit{California Housing} data, latent space search markedly improves plausibility without sacrificing faithfulness: for the \textit{JEM} Ensemble, \textit{ECCCo+} performs substantially better than the baseline and only marginally worse than \textit{REVISE}. Importantly, \textit{ECCCo+} does not attain plausibility at all costs: for the \textit{MLP} Ensemble, plausibility is still very low but this seems to faithfully represent what the model has learned.

We conclude from the findings presented thus far that \textit{ECCCo} enables us to reconcile the objectives of faithfulness and plausibility. It produces plausible counterfactuals if and only if the model itself has learned plausible explanations for the data. It thus avoids the risk of generating plausible but potentially misleading explanations for models that are highly susceptible to implausible explanations.

@@

Another commonly discussed desideratum is closeness~\citep{wachter2017counterfactual}: counterfactuals that are closer to their factuals are associated with smaller costs to individuals in the context of algorithmic recourse. As evident from the additional tables in the appendix, the closeness desideratum tends to be negatively correlated with plausibility and faithfulness. Consequently, both \textit{REVISE} and \textit{ECCCo} generally yield more costly counterfactuals than the baseline. Nonetheless, \textit{ECCCo} does not seem to stretch costs unnecessarily: in Figure~\ref{fig:mnist-bmk} useful parts of the factual `three' are clearly retained.
